---
title: "PCA Notebook"
author: "Simon Ogundare"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

If you've never installed the dependencies listed below on this device before:

```{r install and load dependencies, eval = FALSE}
install.packages("tidyverse")
install.packages("corrr")
install.packages("readxl")
install.packages("ggcorrplot")
install.packages("ggplot2")
install.packages("FactoMineR")
install.packages("factoextra")
```

Load required packages:

```{r}
library(tidyverse)
library(corrr)
library(readxl)
library(ggcorrplot)
library(ggplot2)
library(FactoMineR)
library(factoextra)
```

## Load Data

Load Excel file and normalize numerical data

```{r load CSV with NMR syllable data}
syllable_data <- readxl::read_excel(path = "CSVs/Naked mole-rat colony master sheet (1).xlsx", sheet = 7) %>%
  drop_na(Rank)
#numerical_syllable_data <- syllable_data[, c(5:27)]
numerical_syllable_data <- syllable_data[, c(5:12)]
normalized_numerical_syllable_data <- scale(numerical_syllable_data)
```

Compute and plot correlation matrix using normalized numerical data to look at how well syllable frequency correlates with rank:

```{r look at variable correlations}

correlation_data_frame <- syllable_data[,c(4:12)]
correlation_data_frame <- correlation_data_frame[,c(2:9, 1)]

# Comment this the next two lines with '#' if you do not want to invert rank correlations (if you comment these lines out, high positive correlation indicates that high frequency syllable (red) <=> low ranking). Do not comment these lines out if you do not want to invert rankings:

#########

correlation_data_frame <- arrange(correlation_data_frame, Rank)
correlation_data_frame$Rank <- rev(correlation_data_frame$Rank)

#########

corr_matrix <- cor(correlation_data_frame)
pca_correlations <- ggcorrplot(corr_matrix)

pca_correlations

```

## Principal Component Analysis

Compute PCA and summarize results:

```{r compute pca and summarize}
data.pca <- princomp(corr_matrix)
summary(data.pca)
```

Explore components of interest

```{r explore components 1 and 2}

# interpretation - higher absolute values indicate the factor can explain a larger variance of the sample (a reliable predictor)
# low absolute values for each feature means the factor does not explain an appreciable variance in the sample (a less reliable predictor)

data.pca$loadings[, 1:2]
```

## Visualize principal components

Visualize PCs as Scree plot (visualizes importance of each principal component)

```{r visualize scree plot}
pca_scree <- fviz_eig(data.pca, addlabels = TRUE)
pca_scree
```

Visualize PCA Biplot (visualizes impact of each feature on each principal component)

```{r visualize biplot}
pca_biplot <- fviz_pca_var(data.pca, col.var = "black")
pca_biplot
```

Interpretation (from https://www.datacamp.com/tutorial/pca-analysis-r)

Three main pieces of information can be observed from the previous plot. 

- All the variables that are grouped together are positively correlated to each other.
- The higher the distance between the variable and the origin, the better represented that variable is.
- Variables that are negatively correlated are displayed to the opposite sides of the biplot’s origin.

Visualize Cos2 (contribution of each variable within given component(s))

```{r visualize cos2 plot}
pca_cos2 <- fviz_cos2(data.pca, choice = "var", axes = 1:2)
pca_cos2
```

Visualize Biplot with Cos2 plot

```{r visualize biplot/cos2 plot}
pca_bicos2 <- fviz_pca_var(data.pca, col.var = "cos2",
            gradient.cols = c("black", "orange", "green"),
            repel = TRUE)
pca_bicos2
```

Save visualizations

```{r save visualizations, eval = FALSE}
ggsave(pca_correlations, filename = paste0("PCA_Corr_", Sys.Date(), ".png"), path = "Graphs/")
ggsave(pca_scree, filename = paste0("PCA_Scree_", Sys.Date(), ".png"), path = "Graphs/")
ggsave(pca_biplot, filename = paste0("PCA_Biplot_", Sys.Date(), ".png"), path = "Graphs/")
ggsave(pca_cos2, filename = paste0("PCA_Cos2_", Sys.Date(), ".png"), path = "Graphs/")
ggsave(pca_bicos2, filename = paste0("PCA_Biplot_Cos2_", Sys.Date(), ".png"), path = "Graphs/")
```


